First and foremost, I would like to thank my primary supervisor Jacques Fleuriot.
It is his class that showed me the world of interactive theorem proving and I immediately felt that it fits with how my mind works.
Through working with Jacques, both on my honours project and as part of research internships, I found a love for research.
It was his encouragement that lead me to start on this PhD journey, and it is his support and guidance that saw me through it.
Our discussions always helped me refine my work and filled me with ideas for the future, and I am grateful for all the help he has given me over the years.

In addition to my primary supervisor, I would also like to thank the other members of my supervisory team over the years: Petros Papapanagiotou, Jane Hillston and Ohad Kammar.
Even before I started the PhD, they helped spark and grow my interest in process modelling, both through undergraduate classes and internships.
My work from those internships eventually grew into the research proposal that started my PhD journey.
Throughout these four years, they all provided patient feedback on my research and writing, ensuring I keep growing and improving.
Their help was invaluable.

Throughout my PhD I have been a member of the AI Modelling Lab.
I am deeply grateful to all of them, both for their insightful comments on my work but also the years of friendship.
They took what could have been a cold office and made it into a family with our group lunches, trips to Firbush and celebrations of everyone's achievements.

Finally, I would also like to express my thanks to everyone in my life outside of the School of Informatics.
To my friends, who helped me through moments of self-doubt and were always happy to listen to me talk excitedly about my work.
To my flatmate, who was always there to listen and offer encouragement.
And to my parents, who always supported me and made this all possible.

\medskip
\begin{center}
  \bf
  Thank you all.
\end{center}
